---
title: "Profiling the world's small island developing states"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profiling the world's small island developing states}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5
)
```

The intro vignette shows the mechanics. This one works a small analysis end to
end, using nothing but the bundled `islands` data and the package helpers, to
show the kind of reproducible profile `islandcodes` is meant to make a one-line
job rather than a spreadsheet exercise.

```{r}
library(islandcodes)

sids <- islands[islands$is_sids == 1, ]
nrow(sids)
```

## Who counts, and under what tier

The UN-DESA list mixes fully sovereign states with non-sovereign associate
members. That distinction is analytically load-bearing: an associate member sits
inside a metropolitan constitutional order and inherits a fiscal backstop that a
sovereign micro-state does not. The `sids_tier` column keeps the two apart.

```{r}
table(sids$sids_tier, useNA = "ifany")
```

Overlaying the sub-national island jurisdiction axis shows that the two
classifications are related but not identical. Every SIDS associate member is a
SNIJ, but many SNIJs (Bonaire, the French departments) are not SIDS, because
they are constitutionally integrated into the metropole rather than recognised
as distinct jurisdictions.

```{r}
table(SIDS_tier = sids$sids_tier, SNIJ = sids$is_snij == 1, useNA = "ifany")
```

## How the group distributes across the World Bank taxonomy

Joining the SIDS list onto World Bank region and income group, which is the step
that usually means hand-copying from a PDF, is here just a cross-tab on columns
that already travel with the data.

```{r}
table(sids$wb_region)

table(sids$wb_income_group)
```

The income spread is the point most casual treatments of "SIDS" miss: the group
runs from high-income territories to lower-middle-income states, so any analysis
that treats SIDS as a single development stratum is averaging across a real gap.

## Isolation, measured rather than asserted

"Remote" is asserted about small islands far more often than it is measured.
With coordinates and a distance helper it becomes a computed quantity. For each
SIDS with a known landmass point, take the great-circle distance to its nearest
other SIDS.

```{r}
sids_geo <- sids[!is.na(sids$latitude), ]
codes    <- sids_geo$iso_code

D <- island_distance(codes)          # symmetric matrix of km
diag(D) <- NA                        # ignore self-distance
sids_geo$nearest_sids_km <- round(apply(D, 1, min, na.rm = TRUE))
```

The ranking is a reminder that "distance to the nearest SIDS" measures
clustering, not absolute remoteness. The top of the list mixes oceans: Bahrain
in the Gulf and Singapore in Southeast Asia are not remote in any everyday
sense, yet their nearest fellow SIDS is thousands of kilometres away, and they
sit alongside genuinely oceanic cases such as São Tomé and Príncipe and
Kiribati that are isolated for the opposite reason.

```{r}
ord <- order(sids_geo$nearest_sids_km, decreasing = TRUE)
head(sids_geo[ord, c("label", "wb_region", "nearest_sids_km")], 10)
```

At the other end sit the tightly packed Lesser Antilles, where the nearest peer
is a short hop away: Anguilla and Sint Maarten are 23 km apart.

```{r}
head(sids_geo[order(sids_geo$nearest_sids_km),
              c("label", "wb_region", "nearest_sids_km")], 10)
```

A base-R map makes the two regimes visible at once: the whole reference set in
grey, SIDS overplotted, sized by isolation.

```{r}
plot(islands$longitude, islands$latitude,
     pch = 16, cex = 0.4, col = "grey80",
     xlab = "Longitude", ylab = "Latitude",
     main = "Small island developing states, sized by isolation")
points(sids_geo$longitude, sids_geo$latitude,
       pch = 21, bg = "#f38439", col = "white",
       cex = 0.6 + 2.4 * (sids_geo$nearest_sids_km / max(sids_geo$nearest_sids_km)))
```

## The Dutch Caribbean as a stress test of the coding scheme

The Kingdom of the Netherlands is a compact illustration of why the
disambiguation matters. It comprises seven entities: the European metropole and
six Caribbean territories, which the dataset returns in one filter. They span
both classifications, and the Caribbean six are split geographically into the
ABC islands off Venezuela and the SSS islands some 900 km to the northeast.

```{r}
dc <- islands[islands$political_association == "Dutch Kingdom",
              c("label", "iso_code", "sids_tier", "is_snij")]
dc

# ABC-to-SSS separation, from Aruba
island_distance("AW", c("Sint Maarten", "Saba", "Sint Eustatius"))
```

The Netherlands row earns its place by contrast: it is the only one that is not
a sub-national island jurisdiction and carries no SIDS tier. The three
constituent countries (Aruba, Curaçao, Sint Maarten) are associate-member SIDS,
while Bonaire, Sint Eustatius, and Saba are special municipalities of the
Netherlands proper, so they read as SNIJ but not SIDS. A country-code package
that collapses those three into a single `BQ` row cannot produce this table at
all: half the Caribbean territories would vanish or merge. That is the class of
silent error the package exists to remove, and the reason the analysis above is
reproducible from a single bundled dataset rather than a chain of manual joins.
