---
title: "Profiling the world's small island developing states"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profiling the world's small island developing states}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5
)
```

The intro vignette shows the mechanics. This one works a small analysis end to
end, using nothing but the bundled `islands` data and the package helpers, to
show the kind of reproducible profile `islandcodes` is meant to make a one-line
job rather than a spreadsheet exercise.

```{r}
library(islandcodes)

sids <- islands[islands$is_sids == 1, ]
nrow(sids)
```

## Who counts, and under what tier

The UN-DESA list mixes fully sovereign states with non-sovereign associate
members. That distinction is analytically load-bearing: an associate member sits
inside a metropolitan constitutional order and inherits a fiscal backstop that a
sovereign micro-state does not. The `sids_tier` column keeps the two apart.

```{r}
table(sids$sids_tier, useNA = "ifany")
```

Overlaying the sub-national island jurisdiction axis shows that the two
classifications are related but not identical. Every SIDS associate member is a
SNIJ, but many SNIJs (Bonaire, the French departments) are not SIDS, because
they are constitutionally integrated into the metropole rather than recognised
as distinct jurisdictions.

```{r}
table(SIDS_tier = sids$sids_tier, SNIJ = sids$is_snij == 1, useNA = "ifany")
```

## How the group distributes across the World Bank taxonomy

Joining the SIDS list onto World Bank region and income group, which is the step
that usually means hand-copying from a PDF, is here just a cross-tab on columns
that already travel with the data.

```{r}
table(sids$wb_region)

table(sids$wb_income_group)
```

The income spread is the point most casual treatments of "SIDS" miss: the group
runs from high-income territories to lower-middle-income states, so any analysis
that treats SIDS as a single development stratum is averaging across a real gap.

## Isolation, measured rather than asserted

"Remote" is asserted about small islands far more often than it is measured.
With coordinates and a distance helper it becomes a computed quantity. For each
SIDS with a known landmass point, take the great-circle distance to its nearest
other SIDS.

```{r}
sids_geo <- sids[!is.na(sids$latitude), ]
codes    <- sids_geo$iso_code

D <- island_distance(codes)          # symmetric matrix of km
diag(D) <- NA                        # ignore self-distance
sids_geo$nearest_sids_km <- round(apply(D, 1, min, na.rm = TRUE))
```

The ranking is a reminder that "distance to the nearest SIDS" measures
clustering, not absolute remoteness. The top of the list mixes oceans: Bahrain
in the Gulf and Singapore in Southeast Asia are not remote in any everyday
sense, yet their nearest fellow SIDS is thousands of kilometres away, and they
sit alongside genuinely oceanic cases such as São Tomé and Príncipe and
Kiribati that are isolated for the opposite reason.

```{r}
ord <- order(sids_geo$nearest_sids_km, decreasing = TRUE)
head(sids_geo[ord, c("label", "wb_region", "nearest_sids_km")], 10)
```

At the other end sit the tightly packed Lesser Antilles, where the nearest peer
is a short hop away: Anguilla and Sint Maarten are 23 km apart.

```{r}
head(sids_geo[order(sids_geo$nearest_sids_km),
              c("label", "wb_region", "nearest_sids_km")], 10)
```

A base-R map makes the two regimes visible at once: the whole reference set in
grey, SIDS overplotted, sized by isolation.

```{r}
plot(islands$longitude, islands$latitude,
     pch = 16, cex = 0.4, col = "grey80",
     xlab = "Longitude", ylab = "Latitude",
     main = "Small island developing states, sized by isolation")
points(sids_geo$longitude, sids_geo$latitude,
       pch = 21, bg = "#f38439", col = "white",
       cex = 0.6 + 2.4 * (sids_geo$nearest_sids_km / max(sids_geo$nearest_sids_km)))
```

## From nearest neighbour to access

Distance to the nearest peer is a blunt instrument. It throws away every
observation but one, so an island with a single close neighbour and nothing
else within an ocean scores identically to one sitting inside a dense
archipelago. The economic geography literature, and the recent remote-sensing
work on rural connectivity that borrows from it, uses a decayed sum over all
destinations instead: access is mass discounted by distance, summed.

`island_market_access()` implements that. The package deliberately does not
ship the mass. Population and output are time-varying and contested, so the
figures and their reference year stay the analyst's responsibility. With unit
mass the function reduces to a pure peer-accessibility index, which needs no
external data and isolates what the functional form alone contributes.

```{r}
sids_geo$access <- island_market_access(
  codes,
  mass  = rep(1, length(codes)),   # unit mass: proximity to peers, nothing else
  theta = 1
)
```

The ranking is Caribbean from top to bottom. All twenty highest-access SIDS sit
in the Lesser and Greater Antilles, which is the densest concentration of
island jurisdictions anywhere on the planet.

```{r}
ord_a <- order(sids_geo$access, decreasing = TRUE)
head(sids_geo[ord_a, c("label", "wb_region", "nearest_sids_km", "access")], 10)

table(sids_geo$wb_region[ord_a][1:20])
```

Set against the nearest-neighbour column it is close to a mirror image, and the
Spearman correlation says so. That is worth stating plainly rather than dressing
up: at this level the two measures mostly agree, and access does not overturn
the isolation ranking.

```{r}
cor(sids_geo$nearest_sids_km, sids_geo$access, method = "spearman")
```

What access buys is not a different answer but a better-behaved one.
Nearest-neighbour distance is a single-observation statistic: it is determined
entirely by one peer, so it jumps discontinuously when that peer is added to or
dropped from the reference set, and it cannot distinguish Sint Maarten, which
has one neighbour at 23 km and thirty more within 1,000 km, from a pair of
islands alone in an ocean. Access uses every destination and moves smoothly.

The decay elasticity is a modelling choice, not a fact, so it needs a
sensitivity check rather than a silent default. Here the ordering turns out to
be robust: rank correlation between `theta = 0.5` and `theta = 2` is above 0.97,
and only six of the fifty-eight territories move more than five places.

```{r}
r <- sapply(c(0.5, 1, 2), function(th) {
  rank(-island_market_access(codes, rep(1, length(codes)), theta = th))
})
colnames(r) <- c("theta_0.5", "theta_1", "theta_2")
cor(r[, "theta_0.5"], r[, "theta_2"], method = "spearman")
```

The movement that does happen is concentrated in one place, and it is
substantive rather than numerical noise. Guam and the Northern Mariana Islands
climb fifteen and sixteen places as `theta` rises, because they are a tight pair
inside an otherwise empty stretch of the western Pacific. That is exactly what a
decay parameter encodes: how much a close cluster compensates for global
remoteness. An analyst who reports one ranking under one unstated `theta` has
hidden that judgement rather than made it.

```{r}
moved <- data.frame(label = sids_geo$label, r,
                    move = abs(r[, "theta_0.5"] - r[, "theta_2"]))
head(moved[order(moved$move, decreasing = TRUE), ], 6)
```

With real mass the measure changes character, because access stops being about
how many neighbours there are and becomes about who they are. Here the Dutch
Caribbean six, using illustrative population in thousands: substitute sourced
figures and cite the year in real work.

```{r}
pop <- c(AW = 108, CW = 156, SX = 43, "BQ-BO" = 25, "BQ-SE" = 3, "BQ-SA" = 2)
sort(round(island_market_access(names(pop), pop), 3), decreasing = TRUE)
```

Bonaire tops the within-Kingdom ranking because it sits between the two
population centres of the ABC group, and Sint Maarten comes last despite being
the busiest of the SSS islands, because its only near neighbours are the two
smallest territories in the set. The spread is a factor of seven from top to
bottom.

Widen the destination set to the surrounding mainland economies and that
internal structure almost disappears. The six collapse into two flat tiers, ABC
and SSS, about twenty per cent apart, and within each tier the differences are
negligible.

```{r}
region <- c(US = 335000, CO = 52000, VE = 28000, DO = 11000, PR = 3200)
sort(round(island_market_access(names(pop), region, to = names(region)), 2),
     decreasing = TRUE)
```

The two tables answer different questions. Relative position inside the Kingdom
is governed by which islands you are next to; access to the world outside it is
governed only by which side of the Caribbean you sit on, and the intra-group
distances that dominate the first table are too small to matter in the second.

## What these distances are, and what they are not

Every number above rests on one great-circle distance between two points, and
that construction has limits worth stating before anyone builds on it.

The coordinates are a single representative point per territory. For a compact
island that is unambiguous. For a dispersed archipelago it is not: French
Polynesia spans some 2,000 km of ocean, and a representative point for it is a
convention, not a location. Where the distinction matters, `which = "capital"`
gives the alternative anchor, and the two can differ by hundreds of kilometres.

```{r}
island_coords("French Polynesia")
island_coords("French Polynesia", which = "capital")
```

Great-circle distance is not travel distance and is emphatically not travel
cost. Between two islands there is no road; there is a ferry schedule, an air
route with a hub in between, and a container rotation, each with its own
geometry. Aruba to Sint Maarten is about 900 km as the crow flies and a
connection through a third country in practice. The same caution applies to the
market-access numbers: they are a geographic accessibility measure, and calling
them market access in the economic sense requires the reader to accept distance
as a stand-in for cost. Where actual travel times or freight rates exist, join
them onto these codes and use those instead. That is what the package is for.

Nor does a single point capture internal geography. Two territories of equal
area and equal distance from everything else can differ sharply in how their
own population is distributed relative to a port, which is exactly the variation
the satellite-based literature on rural market access is built to measure and
which a centroid cannot see.

Finally, a handful of territories carry no coordinate at all and propagate as
`NA` rather than being silently dropped: the representative point is missing for
the United States Minor Outlying Islands, and capitals are missing for the
uninhabited and capital-less entries. Checking for those is a step, not an
afterthought.

```{r}
sum(is.na(islands$latitude))
sum(is.na(islands$capital_latitude))
```

## The Dutch Caribbean as a stress test of the coding scheme

The Kingdom of the Netherlands is a compact illustration of why the
disambiguation matters. It comprises seven entities: the European metropole and
six Caribbean territories, which the dataset returns in one filter. They span
both classifications, and the Caribbean six are split geographically into the
ABC islands off Venezuela and the SSS islands some 900 km to the northeast.

```{r}
dc <- islands[islands$political_association == "Dutch Kingdom",
              c("label", "iso_code", "sids_tier", "is_snij")]
dc

# ABC-to-SSS separation, from Aruba
island_distance("AW", c("Sint Maarten", "Saba", "Sint Eustatius"))
```

The Netherlands row earns its place by contrast: it is the only one that is not
a sub-national island jurisdiction and carries no SIDS tier. The three
constituent countries (Aruba, Curaçao, Sint Maarten) are associate-member SIDS,
while Bonaire, Sint Eustatius, and Saba are special municipalities of the
Netherlands proper, so they read as SNIJ but not SIDS. A country-code package
that collapses those three into a single `BQ` row cannot produce this table at
all: half the Caribbean territories would vanish or merge. That is the class of
silent error the package exists to remove, and the reason the analysis above is
reproducible from a single bundled dataset rather than a chain of manual joins.
